A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2019-07-02 doc/large_strain.tex

The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in Section \ref{sec: Constitutive modelling} were implemented using \texttt{deal.II} version 9.0 \cite{dealII90}, an open-source finite element library.
This library offers a number of paradigms by which to parallelize a finite-element code, and as a member of the xSDK (Extreme-scale Scientific Software Development Kit) ecosystem \citep{Bartlett2017b} aims to support exascale computing.
In this work we employ \texttt{deal.II}'s interface to \texttt{p4est} \cite{p4est} library to perform a standard domain decomposition, whereafter each MPI process ``owns'' only a subset of elements and the mesh is distributed across all MPI processes.
%
\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.8\textwidth}
    \centering
    \includegraphics[width=\textwidth]{fine_level_2d.png}
  \end{subfigure}
  \caption{2D mesh after two global refinements distributed into three MPI processes. The color indicates the process owning the element.}%
  \label{fig:miehe_fine_level}
\end{figure}
%
%
Figure \ref{fig:miehe_fine_level} illustrates a distributed MPI decomposition on the fine mesh level for the 2D problem that will be introduced later in Section \ref{sec:example}.

In practice, this implies that parallelization of the assembly operation (for the matrix-based approach) and the linear solver (as well as various pre- and post-processing steps), both of which are the focal points of investigation in this manuscript, has been achieved using MPI.
For the linear solver, \texttt{deal.II}'s implementation of the preconditioned conjugate gradient (CG) solver for symmetric positive-definite systems is consistently used, leaving the choice for method of assembly and linear system preconditioning remaining.

The discrete tangent operator described by \eqref{eq:algebraic_tangent} can be implemented in one of two ways, namely through a classical matrix-based approach or, alternatively, a matrix-free approach.
For the matrix-based approach adopted in this study, the Trilinos \cite{Heroux2005} library was leveraged;
the system tangent matrix is stored in a distributed \textit{Epetra\_FECrsMatrix} from the Epetra package \cite{Heroux2005b} sparse data structure and the linear solver was preconditioned with an algebraic multigrid (AMG) preconditioner from the ML package \cite{Gee2006a}.
This is the most highly performant matrix-based approach using the Trilinos framework that, to date, is possible in \texttt{deal.II}.

@@

The other parts of the GMG preconditioner such as level smoothers, matrix-free interlevel transfer operators and the coarse level iterative solver are provided by the \texttt{deal.II} library, implementational details are discussed in \cite{Clevenger2018}.
%
The efficiency of the GMG preconditioner with the considered level operators will be assessed by numerical examples in the next section.
We will, however, refrain from studying the machine performance aspects of the multigrid preconditioner as a whole as the focus of this paper is the matrix-free implementation of tangent operators of finite-strain elasticity both at the fine mesh level as well as the coarser multigrid levels.
%
Finally, we note that at the time of writing, a $p-$ version of the GMG was not available within the \texttt{deal.II} library and therefore not considered here.
%
% For examples of other approaches that combine homogenization and multigrid methods, see \cite{Bayreuther2003,Fish1995} for applications using small strain elastic materials and \cite{Kaczmarczyk2010} with applications in fracture.

>>>>

The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in Section \ref{sec: Constitutive modelling}, were implemented using \texttt{deal.II} version 9.0 \cite{dealII90}, an open-source finite element library.
This library offers a number of paradigms by which to parallelize a finite-element code, and as a member of the xSDK (Extreme-scale Scientific Software Development Kit) ecosystem \citep{Bartlett2017b} aims to support exascale computing.
In this work we employ \texttt{deal.II}'s interface to the \texttt{p4est} \cite{p4est} library to perform a standard domain decomposition, whereafter each MPI process ``owns'' only a subset of elements and the mesh is distributed across all MPI processes.
%
\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.8\textwidth}
    \centering
    \includegraphics[width=\textwidth]{fine_level_2d.png}
  \end{subfigure}
  \caption{2D mesh after two global refinements distributed into three MPI processes. The color indicates the process owning the element.}%
  \label{fig:miehe_fine_level}
\end{figure}
%
%
Figure \ref{fig:miehe_fine_level} illustrates a distributed MPI decomposition on the fine mesh level for the 2D problem that will be introduced later in Section \ref{sec:example}.

In practice, this implies that parallelization of the assembly operation (for the matrix-based approach) and the linear solver, both of which are the focal points of investigation in this manuscript, has been achieved using MPI;
various pre- and post-processing steps also leveraged parallelization.
For the linear solver, \texttt{deal.II}'s implementation of the preconditioned conjugate gradient (CG) solver for symmetric positive-definite systems is consistently used, leaving the choice for method of assembly and linear system preconditioning remaining.

The discrete tangent operator described by \eqref{eq:algebraic_tangent} can be implemented in one of two ways, namely through a classical matrix-based approach or, alternatively, a matrix-free approach.
For the matrix-based approach adopted in this study, the Trilinos \cite{Heroux2005} library was leveraged;
the system tangent matrix is stored in a distributed \textit{Epetra\_FECrsMatrix} from the Epetra package \cite{Heroux2005b} sparse data structure and the linear solver was preconditioned with an algebraic multigrid (AMG) preconditioner from the ML package \cite{Gee2006a}.
This is the most highly performant matrix-based approach using the Trilinos framework that, to date, is available in \texttt{deal.II}.

@@

The other parts of the GMG preconditioner such as level smoothers, matrix-free interlevel transfer operators and the coarse level iterative solver are provided by the \texttt{deal.II} library; implementational details are discussed in \cite{Clevenger2018}.
%
The efficiency of the GMG preconditioner with the considered level operators will be assessed by numerical examples in the next section.
We will, however, refrain from studying the machine performance aspects of the multigrid preconditioner as a whole as the focus of this paper is the matrix-free implementation of tangent operators of finite-strain elasticity both at the fine mesh level as well as the coarser multigrid levels.
%
Finally, we note that at the time of writing, a $p-$ version of the GMG was not available within the \texttt{deal.II} library and therefore not considered here.
%
% For examples of other approaches that combine homogenization and multigrid methods, see \cite{Bayreuther2003,Fish1995} for applications using small strain elastic materials and \cite{Kaczmarczyk2010} with applications in fracture.
